---
title: "Explore One Variable"
output: html_notebook
---

* Insert a new code chunk: *Cmd+Option+I*
* Run a code cell: *Cmd+Shift+Enter*
* Preview the notebook HTML file: *Cmd+Shift+K*

This notebook will be used to explore a single variable in a data set using the r programming language. 

```{r}
# set the working directory 
setwd('/Users/spencerbertsch/Desktop/dev/CS/MATH177')

# our file is tab separated, not comma separated 
fb_df <- read.csv('assignment0/demo_data/pseudo_facebook.tsv', sep='\t')

names(fb_df)  # <-- gives us the columns or "names" in the dataframe 
```

## Histogram of Users Birthdays
```{r}
# Let's look at a historgam of some user's birthdays 
library(ggplot2)

# here we look at the column headers (names) to get the column we want to plot 
names(fb_df)

qplot(x = dob_day, data = fb_df) + 
  scale_x_continuous(breaks=1:31) + 
  facet_wrap(~dob_month, ncol = 3)  # <-- facet_wrap allows us to easily create subplots based on difference columns! 
```

### Facet Wrap  - faceting over one variable 
Here we can see a great use case of facet_wrap(), a ggplot2 function that lets us generate many of the same types of plots (here histograms), one for each unique catagory of the chosen facet column. 

*facet_wrap(~variable_of_choice)*

### Facet Grid - faceting over two or more variables 
This is similar to facet_grid(), but facet_grid() takes two variables, one we split over the vertical direction, and one we split over the horizontal direction. 

*facet_grid(vertical_variable~horizontal_variable)*

## Lets create histogram of friend counts
```{r}
qplot(x = friend_count, data = subset(fb_df, !is.na(gender)), binwidth = 25) + 
  scale_x_continuous(limits = c(1, 1000), breaks = seq(0, 1000, 50)) + 
  facet_wrap(~gender, ncol = 2)
```

```{r}
by(fb_df$friend_count, fb_df$gender, summary)
```

## Now lets look at Tenure (how long a user has been on fb)
We can plot the x axis over 365 to see the annual tenure! 
```{r}
qplot(x = tenure/365, data = fb_df, 
      xlab = 'Tenure of FB User (Years)', 
      ylab = 'Count of FB Users', 
      binwidth = 0.25, color = I('black'), fill = I('#099DD9')) + 
  scale_x_continuous(breaks = seq(0, 15, 1), lim = c(0, 7))
```


## Histogram of users by age 

```{r}
qplot(x = age, data = fb_df, 
      xlab = 'Age of FB User (Years)', 
      ylab = 'Count of FB Users', 
      binwidth = 1, color = I('black'), fill = I('#FF9400')) + 
  scale_x_continuous(breaks = seq(0, 120, 10))
```



