---
title: "Lab 4 - QBS 177"
output: html_notebook
---

Spencer Bertsch  
Jan. 2022

------------

Some helpful commands:  
* Insert a new code chunk: *Cmd+Option+I*  
* Run a code cell: *Cmd+Shift+Enter*  
* Preview the notebook HTML file: *Cmd+Shift+K*  



### Imports 
```{r}
library(glue)
library(qqman)
```

First we need to change our current working directory to the correct directory for Assignment 3
```{r}
setwd('/Users/spencerbertsch/Desktop/dev/phd_courses/MATH177/lab4')
getwd()
```

### Import the raw data that was generated from the previous lab 
```{r}
load('lab1.Rdata')
```

We new have a mismatch between the number of response variables and samples, so we can restructure the data to make each one equal to 21.  
```{r}
chn <- (fulldat[,5] + fulldat[,6])/2
ind <- (fulldat[,11] + fulldat[,14])/2
nga <- (fulldat[,25] + fulldat[,8])/2
usa <- (0.777*fulldat[,4] + 0.132*fulldat[,2] + 0.053*(chn + ind + fulldat[,16] + fulldat[,15])/4)/(0.777+0.132+0.053)
fulldat <-fulldat[,c(3,1,5,7,9,12,11,24,15,17,19,25,21,20,22,18,13,23,10,4,16)]
fulldat[,3] <- chn
fulldat[,7] <- ind
fulldat[,12] <- nga
fulldat[,20] <- usa
```

We can now read in the smoking data:
```{r}
yy <- read.delim("smoking_outcome.txt")
```


Here we can explicitly obtain the vector that represents our response variable: the prevelance of smoking in 21 different countries. 
```{r}
y <- yy[,2]
```




```{r}
plong <- corlong <- NULL

date()
for (i in 1:10000){
  if(var(fulldat[i,])){   # remove loci with all minor allele frequency 0
    
    # Step 1: We find the correlation between the i'th sample's data and our response variable 
    corlong[i] <- cor(fulldat[i,], y)
    
    # Step 2: Run a linear regression using only the i'th sample's data as predictors
    fit <- lm(y~fulldat[i,]) 
    
    #Step 3: We output slope from linear regression model 
    plong[i] <- summary(fit)$coef[2,4] 
  }
}
date()
```

This process took 2 seconds to finish, so if we assume that processing 10,000 records takes 2 seconds, then processing 1,099,146 samples would take 219.8 seconds, or about 3.5 minutes. 

\n
We can now try to rewrite the above code without using a for loop so that we can improve the runtime of the algorithm. 
```{r}
n <- 21
# We can now obtain the correlation between the futtdat matrix and the response variable: 
# first we scale the fulldat matrix
temp <- scale(t(fulldat))

y1 <- y/sqrt(var(y))

asd <- y1 %*% temp/(n-1)

```


```{r}
fulldat[2, ]
```


We know we have only 21 samples, so $n$ in these calculations is simply 21. 

After we get a t vector, we can use the p-t function to obtain the p-value: 
pt()


```{r}
manhattan(plong)
```


