---
title: "Lab 5 - QBS 177"
output: html_notebook
---

Spencer Bertsch  
Jan. 2022

------------

Some helpful commands:  
* Insert a new code chunk: *Cmd+Option+I*  
* Run a code cell: *Cmd+Shift+Enter*  
* Preview the notebook HTML file: *Cmd+Shift+K*  



### Imports 
```{r}
library(glue)
```

First we need to change our current working directory to the correct directory for Assignment 3
```{r}
setwd('/Users/spencerbertsch/Desktop/dev/phd_courses/MATH177/lab5')
getwd()
```

### Import the raw data
```{r}
data <- read.table("fulldat.txt", header=F)
data[1:3,1:10]
y <- data[,1]
x <- data[,3:50]
workingD <- data.frame(cbind(y,x))
```
We now want to randomly create the train-test split for our data set 
```{r}
set.seed(1)
train <- sample(1:240,160,replace=FALSE) # randomly divide the data in to training and testing
test <- (1:240)[-train]
```


Lets examine the dimensions of our new training and testing sets :
```{r}
length(train)
length(test)
```


We now want to run a stepwise regression on our training set with 160 samples
```{r}
#define intercept-only model
intercept_only <- lm(mpg ~ 1, data=mtcars)

#define model with all predictors
all <- lm(mpg ~ ., data=mtcars)

#perform forward stepwise regression
forward <- step(intercept_only, direction='forward', scope=formula(all), trace=0)
```









