---
title: "Lab 3 - QBS 177"
output: html_notebook
---

Spencer Bertsch  
Jan. 2022

------------

Some helpful commands:  
* Insert a new code chunk: *Cmd+Option+I*  
* Run a code cell: *Cmd+Shift+Enter*  
* Preview the notebook HTML file: *Cmd+Shift+K*  


### Imports 
```{r}
library(glue)
```

First we need to change our current working directory to the correct directory for lab3
```{r}
setwd('/Users/spencerbertsch/Desktop/dev/phd_courses/MATH177/lab3')
getwd()
```


```{r}
melanoma_df <- read.table("PCA.example.melanoma.txt",header=TRUE)
mu = colMeans(melanoma_df)
```


```{r}
for(i in 1:ncol(melanoma_df)){
  if (i %% 500 == 0){
    print(glue('Processing column {i} and replacing all NULL values with column mean...'))
  }
  melanoma_df[is.na(melanoma_df[,i]), i] <- round(mean(melanoma_df[,i], na.rm = TRUE))
}
```

```{r}
print(glue('We ca now double check that our new melanoma dataframe contains no nulls: {sum(is.na(melanoma_df))}'))

```


Here we can use principal component analysis to transform the initially high dimensional dataframe 
```{r}
melanoma_df.pc.eigen.cor <- prcomp(melanoma_df)
```


Note that the above command 'prcomp' is equivalent to finding the covariance matrix of the initial high-dimensional matrix, then find the eigenvalues of that matrix: 
```
df 
c_df = cov(df)
eighen(c_df)
```

```{r}
names(melanoma_df.pc.eigen.cor)
mel.scores.eigen <- melanoma_df.pc.eigen.cor$x
mel.scores.rotation <- melanoma_df.pc.eigen.cor$rotation
```

```{r}
Xhat = mel.scores.eigen[,1:nComp] %*% t(mel.scores.rotation[,1:nComp])
# Xhat[1,]
```


```{r}
cov_matrix <- cov(Xhat)
```


```{r}
screeplot(melanoma_df.pc.eigen.cor, type = "l", npcs = 15, main = "Screeplot of the first 15 PCs")
abline(h = 1, col="red", lty=5)
legend("topright", legend=c("Eigenvalue = 1"),
col=c("red"), lty=5, cex=0.6)
```


